### 2020-09-16 Manuscript/manuscript.tex
small edits as well as additions to glossary

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}). Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian orogeny collision \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, rifting occurred ca. 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}. However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia). If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. The record of Neoproterozoic basin development lead \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

@@

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca. 700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia). This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys). Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca. 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2019a} are also included.

@@

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia \citep{Irving1972a}. Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop.

@@

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there are inconsistencies between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 583 and 565 Ma (Fig. \ref{fig:Laurentia_paleolatitude}). This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g. \citealp{Pisarevsky2001a,Li2008a}). One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca. 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization and that interpret steeper directions also recovered from the intrusives as the result of remagnetization. They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca. 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization. Notably the lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole. Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Data from the ca. 585 Ma Grenville dyke swarm interpreted as primary reveal $\sim$90\textdegree$\;$differences in direction within dikes dated within to high-precision 4 million years of one another \citep{Halls2015a}. The rates of $>$20\textdegree /Myr ($>$220 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions. Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak which could support a deviation from stable axial dipolar geomagnetic field behavior at the time \citep{Thallner2020a}.  Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

A synthesized pole path for Laurentia is developed here and used to develop a paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. In portions of the record with a more dense record of poles, such as ca. 1450 Ma, a calculated running mean is used to integrate constraints from multiple poles. This method follows the approach taken in the Phanerozoic (e.g. \citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps. When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model. Where there are no constraints a simple interpolation between constraints is made. While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain. These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca. 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}. The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.

@@

Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes with specified paleolongitudes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models. This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
\textit{Note that this section is currently under development and is incomplete}

@@

Could discuss Gibson2019b model related to hypothesized Paleoproterozoic tectonic history

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}). Proterozoic orogens are labeled with \textit{italicized text} (or. -- orogen; THO  -- Trans-Hudson orogen; prov. -- province). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

This late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}. This constraint is incorporated into models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a, Elming2021a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and (an)other continent(s) ca. 1080 Ma -- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, rifting occurred ca. 780 to 720 Ma leading to deposition in basins from the Death Valley region of SW Laurentia to the Mackenzie Mountains of NW Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}. However, this extensional basin development predates the rifting that led to well-documented passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary \citep{Bond1984a, Levy1991a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province along this margin and the subsequent extension recorded in the basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia). If this interpretation is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (post 635 Ma as in \citealp{Levy1991a} and \citealp{Witkosky2018a}). The geological evidence therefore supports prolonged active tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), but suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. The record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a}, is that ca. 780 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further outboard. In this model, subsequent continent rifting that drove lithospheric thinning, perhaps associated with the departure of a microcontinent fragment rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

@@

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca. 700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia). This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys). Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia margin ca. 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results \citep{Evans2021b}. While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland \citep{Brown2018a}. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and associated cooling within the Grenvillian orogen. Cooling ages of deeply exhumed lithologies are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions (Table 2). However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southwest Laurentia (ca. 760 Ma)  as presented in \cite{Eyster2019a} are also included.

@@

Data from other terranes add resolution to the record. In particular, data from Greenland add 12 poles between 1385 and 1160 Ma when there are only four poles from mainland Laurentia. Given that the rotation between Greenland and mainland Laurentia is well-constrained (Table \ref{tab:terrane_rotations}), once rotated these poles can be used for reconstruction of the entire continent. The reliability of this approach gains credence through the good agreement between the ca. 1633 Ma Melville Bugt diabase dikes pole from Greenland \citep{Halls2011a} and the ca. 1590 Ma Western Channel diabase pole of mainland Laurentia (\citealp{Irving1972a}; Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). Similarly, there is good agreement between the ca. 1267 Ma Mackenzie dikes pole of Laurentia \citep{Buchan2000a} and coeval poles from Greenland such as the ca. 1275 Ma North Qoroq intrusives \citep{Piper1992a} and Kungnat Ring dike \citep{Piper1977a}. Furthermore, the Greenland poles with ages that fall between the ca. 1237 Ma Sudbury dikes and ca. 1144 Ma lamprophyre dikes pole of mainland Laurentia are consistent with constraints on either side from the mainland while filling in the ascending limb of the path leading up to the apex of 1140 to 1108 Ma poles known as the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}).

@@

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period at the end of the Neoproterozoic era where there are inconsistencies between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 583 and 565 Ma (Fig. \ref{fig:Laurentia_paleolatitude}). This conflicting record is a longstanding problem and has led to the presentation of both high-latitude and low-latitude Laurentia paleogeographic reconstructions at the time (e.g. \citealp{Pisarevsky2001a,Li2008a}). One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Paleodirectional data from single feldspar crystals from the Sept-\^Iles layered intrusion led \cite{Bono2015a} to interpret the lower inclination (and therefore lower latitude) direction from the intrusion (the one included as the ca. 565 Ma Sept-\^Iles pole in Table 2; \citealp{Tanczyk1987a}) as the primary thermal remanent magnetization.  \cite{Bono2015a} interpret steeper directions also recovered from the intrusives as the result of remagnetization. They suggest that other steep magnetizations from Ediacaran Laurentia plutonic rocks, such as that observed in the ca. 583 Ma Baie des Mountons complex (the A group of \cite{McCausland2011a} in Table 2), are also the result of remagnetization. Notably the lower inclination Baie des Mountons complex B Group directions result in a pole that is indistinguishable from the lower inclination  Sept-\^Iles intrusives pole. Another possibility discussed in the literature is that the lack of congruency between poles in this time interval is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Data from the ca. 585 Ma Grenville dyke swarm of Laurentia interpreted as primary reveal $\sim$90\textdegree$\;$differences in direction within dikes dated within to 2.5 $\pm$ 0.9 million years of one another \citep{Halls2015a}. The rates of $>$26\textdegree /Myr ($>$288 cm/yr) implied if these data are interpreted as resulting from plate motion or true polar wander were considered as dynamically implausible by \cite{Halls2015a} leading the authors to favor a deviation from axial dipolar behavior as the explanation for disparate Ediacaran directions. Estimates of magnetic paleointensity in these Grenville dikes are anomalously weak which could support a deviation from stable axial dipolar geomagnetic field behavior at the time \citep{Thallner2020a}.  Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the compiled record for Laurentia (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

A synthesized pole path for Laurentia is developed here and used to develop a paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. In portions of the record with a more dense record of poles, such as ca. 1450 Ma, a calculated running mean is used to integrate constraints from multiple poles. This method follows the approach taken in the Phanerozoic (e.g. \citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps. When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model. Where there are no constraints a simple interpolation between constraints is made. While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain. These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca. 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}. The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China (albeit one that needs to be revisited given updates to the paleomagnetic and geochronologic record from South China; \citealp{Zhang2021a}).

@@

Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes with specified paleolongitudes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions}, the map projections are centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia. This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
\textit{Note that this section is currently under development and is incomplete}

@@

%Could discuss Gibson2019b model related to hypothesized Paleoproterozoic tectonic history

@@

\noindent\textbf{allochthonous } An adjective denoting that a rock or terrane originated in a position at significant distance from the lithospheric block that it is currently a part of.

@@

\noindent\textbf{conjugate } Adjective referring to continents or continental margins that were previously conjoined.

@@

\noindent\textbf{Elzevirian orogen } The orogen resulting from the Mesoproterozoic Elzevirian orogeny when there was accretion of arc terranes to eastern Laurentia.

\noindent\textbf{evaporite } Minerals that crystallize from water that becomes supersaturated in salts through evaporation.

@@

\noindent\textbf{geocentric axial dipole hypothesis } The hypothesis that when it is time-averaged, Earth's magnetic field is dominantly a dipole aligned with the spin axis.

\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from southerneastern Ontario, Canada to west Texas, USA.

@@

\noindent\textbf{Hadley cell } Large-scale atmospheric circulation where air rises near the equator, flows poleward, and descends in the subtropics. This circulation drives convective tropical precipitation and the dry downwelling air leads to aridity in the subtropics.

@@

\noindent\textbf{hematite } An iron oxide mineral with a formula of Fe\textsubscript{2}O\textsubscript{3} that commonly holds magnetization in geologic materials, particularly oxidized sedimentary rocks.

\noindent\textbf{juvenile } Rocks that have formed through melt recently extracted from the mantle.

@@

\noindent\textbf{Midcontinent Rift } A major Mesoproterozoic intracratonic rift where there was co-location of large igneous province magmatism and extension in Laurentia's interior centered on the Lake Superior region.

@@

\noindent\textbf{paleolatitude } The past latitude of a given point on Earth's surface at a given time.

\noindent\textbf{paleomagnetic pole } A calculated position from paleomagnetic data that is interpreted to correspond to the ancient position of Earth's spin axis (the north pole) through application of the geocentric axial dipole hypothesis. The uncertainty on the pole position is given as a circle with a radius of a given angle ($A_{95}$).

@@

\noindent\textbf{Phanerozoic } A geologic eon spanning from 541 million years ago to the present day.

\noindent\textbf{Picuris orogen } An orogen resulting from a Mesoproterozoic orogeny interpreted from metamorphic rocks with Mesoproterozoic-aged protoliths in northern New Mexico, USA.

@@

\noindent\textbf{Shawinigan orogen } The orogen resulting from the Mesoproterozoic Shawinigan orogeny when there was accretion of terranes to eastern Laurentia.

@@

\noindent\textbf{thermal remanent magnetization } Magnetization acquired by magnetic minerals rocks as they cool following crystallization from magma.
